---
title: "SEISMIC Collaborative LAK Workshop"
author: "V.Farrar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
syndat <- read.csv(file = "SEISMIC_synthetic_data_2023-01-26.csv")

#load packages
library(tidyverse)
```

# Define Your Course of Interest

What course are you interested in looking at? Options: BIO300, MATH100

Define it below, in quotes "", in the code below. This will be used in the rest of the code, and you can always come back and change it.

```{r cars}
course_of_interest <- "MATH100"
```

## Calculating SAI

SAI, or systemic advantage index, is a metric that takes into account multiple axes of student identities, including:

-   race/ethnicity

-   gender

-   socioeconomic status

-   parental education (i.e., first-generation college-going status)

**Your task: Walk through one of the lines of code and compare it with the SAI table below from the talk. Can you explain where the SAI assignment comes from?**

```{r SAI calculation, echo=TRUE}
#this code goes through all possible combinations of the 4 advantage axes and assigns an SAI based on that students' values
#as ethniccode_cat == 1 is the "BIPOC" label, all other categories are considered advantaged 
syndat <-
  syndat %>% mutate(sai = case_when(
    female == "0" & ethniccode_cat != "1" & firstgen == "0" & lowincomeflag == "0"  ~ "4",
    female == "1" & ethniccode_cat != "1" & firstgen == "0" & lowincomeflag == "0"  ~ "3",
    female == "0" & ethniccode_cat == "1" & firstgen == "0" & lowincomeflag == "0"  ~ "3", 
    female == "0" & ethniccode_cat != "1" & firstgen == "0" & lowincomeflag == "1"  ~ "3", 
    female == "0" & ethniccode_cat != "1" & firstgen == "1" & lowincomeflag == "0"  ~ "3", 
    female == "0" & ethniccode_cat == "1" & firstgen == "1" & lowincomeflag == "0"  ~ "2", 
    female == "0" & ethniccode_cat == "1" & firstgen == "0" & lowincomeflag == "1"  ~ "2", 
    female == "1" & ethniccode_cat == "1" & firstgen == "0" & lowincomeflag == "0"  ~ "2", 
    female == "0" & ethniccode_cat != "1" & firstgen == "1" & lowincomeflag == "1"  ~ "2", 
    female == "1" & ethniccode_cat != "1" & firstgen == "1" & lowincomeflag == "0"  ~ "2", 
    female == "1" & ethniccode_cat != "1" & firstgen == "0" & lowincomeflag == "1"  ~ "2", 
    female == "1" & ethniccode_cat != "1" & firstgen == "1" & lowincomeflag == "1"  ~ "1", 
    female == "1" & ethniccode_cat == "1" & firstgen == "1" & lowincomeflag == "0"  ~ "1",
    female == "1" & ethniccode_cat == "1" & firstgen == "0" & lowincomeflag == "1"  ~ "1", 
    female == "0" & ethniccode_cat == "1" & firstgen == "1" & lowincomeflag == "1"  ~ "1",
    female == "1" & ethniccode_cat == "1" & firstgen == "1" & lowincomeflag == "1"  ~ "0",
    TRUE ~ "NA"))
```

Let's explore what grade outcomes look like across levels of SAI in our course of interest.

```{r SAI summary stats, warning=FALSE, message=FALSE}
syndat %>%
  filter(crs_name == course_of_interest) %>%
  group_by(crs_name, sai) %>%
  summarise(n = n(),
            mean_grade = mean(numgrade, na.rm = T),
            sd_grade = sd(numgrade, na.rm = T))
```
